Gradual Soundness: Lessons from Static Python

### 2021-10-26 ecoop-2022/paper.tex
ecoop: ctypes overview

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific instructions for the Cinder runtime system.
The type system can express common Python idioms and is guaranteed
sound through run-time checks, which in turn enable optimizations.
The type system also includes a dynamic type as an escape hatch to
untyped behavior.
An unannotated variable behaves exactly as it would in Python.

@@

\subsubsection{CTypes}

[FILL]
CTypes for primitives, unboxed representation.
Casts and unboxes happen at the boundary to Python.
Straightforward, not included in model.
Important for optimizations.

@@

@@

>>>>

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific instructions for the Cinder runtime.
The type system can express common Python idioms and is guaranteed
sound through run-time checks, which in turn enable optimizations.
The type system also includes a dynamic type as an escape hatch to
untyped behavior.
An unannotated variable behaves exactly as it would in Python.

@@

@@

  \item
    \textbf{Checked Dict}
    Implementation too, to support the types.

@@

\subsubsection{Primitive Types}

For performance-critical applications, \SP{} includes a \emph{distinct} set of primitive types.
The basic primitives describe booleans and sized numbers, e.g., \code{int64}, \code{uint64}, \code{double}.
Two additional primitives, \code{Array} and \code{Vector}, describe collections.
Cinder implements these types with unboxed C numbers and C arrays, which cost
(and do!) much less than their Python counterparts.

% For now, programmers have to write and manage primitive types.
% In the future, a preprocessor might convert Python arithmetic to primitive arithmetic.

Primitives are carefully limited by the type system.
Neither a module-level nor a closure-level variable may have a primitive
type---because untyped code can mutate such variables.
Primitive types are incompatible with the dynamic type,
and there are no implicit upcasts from primitives
to a matching Python type.
Even a conjunction disallows mixing; for example, the \code{and} operator
requires either two Python booleans or two primitive booleans.

The only way for Python values and primitive values to interact
is through dedicated conversion functions.
Cinder handles conversions at the boundaries between \SP{} code
and untyped code to avoid a cascade of modify-then-run refactorings, but nowhere else.
Within typed code, programmers must satisfy the type checker with
appropriate conversions.
